A Systematic Literature Review on Extensions to Role-Based Access Control

### 2014-02-11 paper/sections/abstract.tex
Updates for introduction references and research question refining.
Updating counts in appendix and other data tables.

\noindent
{\bf Context}: Since the United States National Institute of Standards and Technology proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-targeted extensions have been proposed.  For example, the mobile domain has surfaced a need for permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how RBAC extension models are evaluated by providing an assessment of the state of RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review of RBAC extension models that began with 1,716 papers of which 27 were deemed as primary sources for inclusion.
\\
\\
\noindent
{\bf Results}: We identified and classified the RBAC extension models into eight extension categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. Only 8 of the 27 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains 
that inspired extensions were the medical domain with 9 of the 27 models, enterprise workflows with 5 of 27, and mobile computing with 5 of 27.
\\
\\
\noindent
{\bf Conclusions}: Our literature review shows that all eight of the RBAC extension categories we identified deal with context. The state of evaluation of RBAC model extensions is limited with 1 paper doing an evaluation comparing itself to the RBAC Reference Model, 8 presenting example 
scenarios of their model in action, and 12 of the 27 models lacking an evaluation that we could discern. The magnitude and scope of extensions to the RBAC standard suggests a revised standard may be beneficial.

>>>>

\noindent
{\bf Context}: Since the United States National Institute of Standards and Technology proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-targeted extensions have been proposed.  For example, the mobile domain has identified a need for permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding how RBAC extension models are evaluated by providing an assessment of the state of RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review of RBAC extension models that began with 1,716 papers of which 27 were deemed as primary sources for inclusion.
\\
\\
\noindent
{\bf Results}: We identified and classified the RBAC extension models into eight extension categories: Constraint, Context, Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. Only 8 of the 27 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains that inspired extensions were the medical domain with 9 of the 27 models, enterprise workflows with 5 of 27, and mobile computing with 5 of 27. The state of evaluation of RBAC model extensions is inconsistent with 4 papers comparing themselves to the RBAC Reference Model, 14 presenting example scenarios, and 12 lacking any evaluation that we could discern. 
\\
\\
\noindent
{\bf Conclusions}: Our literature review shows that all eight of the RBAC extension categories we identified deal with context whereby the privileges provided to a role are environmental depending upon factors such as location or time. 
The magnitude and scope of extensions to the RBAC standard suggests a revised standard may be beneficial.

### 2014-02-11 paper/sections/analysis.tex
Updating counts in appendix and other data tables.

\section{RQ3: Evaluations} \label{sec:evaluations}

@@

When designing and proposing a model targeted at a feature that is rooted in practical usage by real software systems, bringing the model to life is strong evidence that the proposed model can work in practice. 
The concept of authorization and access control is rooted in a business need. 
Thus, any access control model needs to be feasible in the real world not just on paper. 
We analyzed the primary sources to see how many proposed models actually had implementations associated with them. 
Then, we quantified types of implementations. 
Whether the implementation was for a real system, for a prototype and/or used in a production environment.

The 27 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors.
Further, we identified a set of evaluation types found within  each of the primary sources and provide below a list of the 
evaluation types and which primary sources provided which type.  In some cases a single primary source provided multiple evaluation
types.

@@

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 27 papers surveyed, 4 papers~\cite{zhou2007team,bao08:role, cholewka00:acontext-sensitive,huang06:pervasive} were simply prototypes developed by the authors whereas 4 papers~\cite{jian2008extended, aich09:role, yao2008task, motta03:contextual} were claimed to be implemented within a real
system. The remaining 20 papers provide no
mention of an implementation or prototype.

Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. 14 models presented example scenarios
and how application of their model would apply and resolve the situation.  6 models provided some form of performance
or complexity analysis of their model.  The performance and complexity analysis included graphs of the model's time to determine calculate authorization 
as the number of entities grew, and the size of the role space for the extension model compared to the RBAC Reference Model. 
5 models provided mathematical descriptions and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation approach was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
or show how the Reference Model is deficient in tackling said scenarios.

@@

The RBAC standard is designed with enterprises in mind such that when practitioners implemented RBAC into their systems,
there would be a reasonable assurance being based off a well thought out model.  As extensions to the RBAC Reference Mode
come along, thought and time should be given to how features and nuances of their models may impact implementation
in order to achieve the same goals as the original standard.  The primary sources should a significant lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as four papers did, should be
a prototype implementation of the model for review by both practitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

\textbf{Implementation Type} & \textbf{Papers} & \textbf{Count} \\ \cline{1-3}
Enterprise
&
\cite{motta03:contextual}
\cite{aich09:role}
\cite{yao2008task}
&
3 \\ \cline{1-3}
Prototype
& 
\cite{cholewka00:acontext-sensitive}
\cite{huang06:pervasive}
\cite{bao08:role}
\cite{zhou07:network}
&
4 \\ \cline{1-3}
None
&
\cite{alam06:constraint}
\cite{tzelepi01:flexible}
\cite{yamazaki04:designing}
\cite{jian2008extended}
\cite{thein2011leveraging}
\cite{zou2009crbac}
\cite{haibo05:context}
\cite{zhang06:collaborative}
\cite{ni2010privacy}
\cite{masoumzadeh2008purbac}
\cite{damiani2007geo}
\cite{ray2006lrbac}
\cite{hansen2003spatial}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
\cite{samuel07:spatio-temporal}
\cite{chandran05:llt}
\cite{ray07:spatio}
\cite{oh2003task}
\cite{joshi05:generalized}
&
20 \\

@@

\greybox{Research on RBAC extension models shows a lack of application and testing in the real world scenarios that the models are ultimately designed for. 
Further, comprehensive evaluation of the models were found to be significantly lacking. }

@@

\textbf{Evaluation Type} & \textbf{Papers} & \textbf{Count} \\ \cline{1-3}
Time-Based Performance
&
\cite{ni2010privacy}
\cite{aich09:role}
&
2 \\ \cline{1-3}
Complexity Analysis
&
\cite{bao08:role}
\cite{zhang06:collaborative}
\cite{chen08:spatio-temporal}
\cite{aich09:role}
&
4 \\ \cline{1-3}
Comparison to Standard RBAC
& 
\cite{bao08:role}
\cite{zou2009crbac}
\cite{zhang06:collaborative}
\cite{zhao2008flexible}
\cite{ray07:spatio}
&
5 \\ \cline{1-3}
Mathematical Modeling
& 
\cite{damiani2007geo}
\cite{hansen2003spatial}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
\cite{joshi05:generalized}
&
5 \\ \cline{1-3}
Example Scenarios of the Model in Action
& 
\cite{alam06:constraint}
\cite{tzelepi01:flexible}
\cite{cholewka00:acontext-sensitive}
\cite{huang06:pervasive}
\cite{bao08:role}
\cite{jian2008extended}
\cite{yamazaki04:designing}
\cite{zou2009crbac}
\cite{ray07:spatio}
\cite{samuel07:spatio-temporal}
\cite{joshi05:generalized}
\cite{yao2008task}
\cite{zhou2007team}
\cite{oh2003task}
&
14 \\ \cline{1-3}
Case Study of the Model in Practice
&
\cite{motta03:contextual}
&
1 \\ \cline{1-3}

\end{tabular*}
\label{tab:implementations}
\end{table}

\section{RQ4: Domains} \label{sec:domains}

\textit{What domains have RBAC extensions been created for?}
\\

@@

\section{RQ5: Generalizations} \label{sec:generalizations}

\textit{What commonalities or generalizations exist across all categories?}
\\

During the paper reading phase, we identified commonalities or generalizations within the primary sources by looking for formal modeling of extending RBAC extension models.

>>>>

\section{RQ3: Implementations and Evaluations} \label{sec:implementations}

@@

When designing and proposing a model targeted at a feature that is rooted in practical
usage by real software systems, bringing the model to life is evidence that the
proposed model can work in practice. The presence of an actual implementation substantiates the robustness of the design and/or the need for the extension.
We analyzed the primary sources to see how many
proposed models actually had implementations associated with them.
We quantified whether the implementation was enterprise (i.e., an implementation used in a production environment) was for a real system) or a prototype implementation (i.e., proof of concept to demonstrate the feasibility).

Furthermore, the 27 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors. We identified a set of evaluation types found within  each of the primary sources and provide a list of the evaluation types and which primary sources provided which type.
In some cases a single primary source provided multiple evaluation types.

@@

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 27 papers surveyed, 4 papers~\cite{zhou2007team,bao08:role, cholewka00:acontext-sensitive,huang06:pervasive} were prototypes developed by the authors whereas 4 papers~\cite{jian2008extended, aich09:role, yao2008task, motta03:contextual} were claimed to be implemented within a real
system. The remaining 20 papers provide no
mention of an implementation.

Table \ref{tab:evaluations} shows evaluation types.
Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. 14 models presented example scenarios
and how application of their model would apply and resolve the situation. Six models provided some form of performance
or complexity analysis of their model.  The performance and complexity analysis included graphs of the model's time to determine calculate authorization 
as the number of entities grew, and the size of the role space for the extension model compared to the RBAC Reference Model. 
Five models provided mathematical descriptions and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation approach was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
or show how the Reference Model is deficient in tackling said scenarios.

\textbf{Implementation Type} & \textbf{Papers} & \textbf{Count} \\ \cline{1-3}
Enterprise
&
\cite{motta03:contextual}
\cite{aich09:role}
\cite{yao2008task}
&
3 \\ \cline{1-3}
Prototype
& 
\cite{cholewka00:acontext-sensitive}
\cite{huang06:pervasive}
\cite{bao08:role}
\cite{zhou2007team}
&
4 \\ \cline{1-3}
None
&
\cite{alam06:constraint}
\cite{tzelepi01:flexible}
\cite{yamazaki04:designing}
\cite{jian2008extended}
\cite{thein2011leveraging}
\cite{zou2009crbac}
\cite{haibo05:context}
\cite{zhang06:collaborative}
\cite{ni2010privacy}
\cite{masoumzadeh2008purbac}
\cite{damiani2007geo}
\cite{ray2006lrbac}
\cite{hansen2003spatial}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
\cite{samuel07:spatio-temporal}
\cite{chandran05:llt}
\cite{ray07:spatio}
\cite{oh2003task}
\cite{joshi05:generalized}
&
20 \\

@@

The RBAC standard is designed in mind such that when practitioners implemented RBAC into their systems, The RBAC standard demonstrates reasonable assurance being based off a well thought out model.
As extensions to the RBAC Reference Mode come along, thought and time should be given to how features and nuances of their models may impact implementation to achieve the same goals as the original standard. 
The primary sources showed a lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as four papers did, should be
a prototype implementation of the model for review by both practitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

@@

\greybox{Research on RBAC extension models shows a lack of implementations in the real world scenarios that the models are ultimately designed for. RBAC extension model research lacks comprehensive evaluation of the models both in theory and in practice.}

@@

\textbf{Evaluation Type} & \textbf{Papers} & \textbf{Count} \\ \cline{1-3}
Time-Based Performance
&
\cite{ni2010privacy}
\cite{aich09:role}
&
2 \\ \cline{1-3}
Complexity Analysis
&
\cite{bao08:role}
\cite{zhang06:collaborative}
\cite{chen08:spatio-temporal}
\citep{aich07:STARBAC}09:role}
&
4 \\ \cline{1-3}
Comparison to Standard RBAC
& 
\cite{bao08:role}
\cite{zou2009crbac}
\cite{zhang06:collaborative}
\cite{zhao2008flexible}
\cite{ray07:spatio}
&
5 \\ \cline{1-3}
Mathematical Modeling
& 
\cite{damiani2007geo}
\cite{hansen2003spatial}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
\cite{joshi05:generalized}
&
5 \\ \cline{1-3}
Example Scenarios of the Model in Action
& 
\cite{alam06:constraint}
\cite{tzelepi01:flexible}
\cite{cholewka00:acontext-sensitive}
\cite{huang06:pervasive}
\cite{bao08:role}
\cite{jian2008extended}
\cite{yamazaki04:designing}
\cite{zou2009crbac}
\cite{ray07:spatio}
\cite{samuel07:spatio-temporal}
\cite{joshi05:generalized}
\cite{yao2008task}
\cite{zhou2007team}
\cite{oh2003task}
&
14 \\ \cline{1-3}
Case Study of the Model in Practice
&
\cite{motta03:contextual}
&
1 \\ \cline{1-3}
None
&
\cite{masoumzadeh2008purbac}
\cite{chandran05:llt}
\cite{ray2006lrbac}
&
3 \\ \cline{1-3}

\end{tabular*}
\label{tab:evaluations}
\end{table}

\section{RQ5: Domains} \label{sec:domains}

\textit{What domains have been targeted by RBAC extension models?}
\\

@@

\section{RQ6: Generalizations} \label{sec:generalizations}

\textit{What commonalities exist across RBAC extension models?}
\\

During the paper reading phase, we identified commonalities within the primary sources
by looking for formal modeling of extending RBAC extension models.

### 2014-02-11 paper/sections/appendix.tex
Adding back analysis section and cleaning up.
Updating counts in appendix and other data tables.

\begin{table}
\centering
\caption{Primary sources by classification and properties (Implementation Type / Domain / Evaluation Type(s))}
\vline
\begin{tabularx}{\textwidth}{ | c | c | c | c | l | l | }
\cline{1-4}

@@

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ | p{.95\textwidth} | }{\cite{haibo05:context} A Context-Aware Role-Based Access Control Model for Web Services, 2005 (None / Web / None)} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ | p{.95\textwidth} | }{\cite{jian2008extended} An extended RBAC model based on granular logic, 2008 (None / Organizations / Scenarios)} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{1}{*}{\rotatebox{90}{Const. }}}} &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{alam06:constraint} A constraint based role based access control in the SECTET a model-driven approach, 2006 } \\[4ex] \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{1}{*}{\rotatebox{90}{ Org. }}}} &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{zhang06:collaborative} ROBAC: Scalable Role and Organization Based Access Control Models, 2006 } \\[2.75ex]  \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{2}{*}{\rotatebox{90}{Priv. }}}} &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{ni2010privacy} Privacy-aware role-based access control, 2007 } \\[1ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{masoumzadeh2008purbac} PuRBAC: Purpose-Aware Role-Based Access Control, 2008 } \\[1ex] \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Task }}}} &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{yao2008task} A Task-Role Based Access Control Model with Multi-Constraints, 2008 } \\ \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{zhou2007team} Team and Task Based RBAC Access Control Model, 2009 } \\ \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{oh2003task} Task-role-based access control model, 2003 } \\ \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{10}{*}{\rotatebox{90}{ Spatio-Temporal }}}} &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{aich07:STARBAC} STARBAC: Spatiotemporal role based access control, 2007 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{chen08:spatio-temporal} On spatio-temporal constraints and inheritance in role-based access control, 2008 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{samuel07:spatio-temporal} A framework for specification and verification of generalized spatio-temporal role based access control model, 2007 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{chandran05:llt} LoT-RBAC: A Location and Time-Based RBAC Model, 2005 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{ray07:spatio} A Spatio-temporal Role-Based Access Control Model, 2007 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.7\textwidth} | }{\cite{aich09:role} Role Based Access Control with Spatiotemporal Context for Mobile Applications, 2009 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Spatial }}}} &
\multicolumn{1}{ | p{.6\textwidth} | }{\cite{damiani2007geo} GEO-RBAC: a spatially aware RBAC, 2005 } \\[0.25ex] \cline{4-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{}  &
\multicolumn{1}{ | p{.6\textwidth} | }{\cite{ray2006lrbac} LRBAC: A location-aware role-based access control model, 2006 } \\[0.25ex] \cline{4-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{}  &
\multicolumn{1}{ | p{.6\textwidth} | }{\cite{hansen2003spatial} Spatial role-based access control model for wireless networks, 2003 } \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Temp. }}}} &
\multicolumn{1}{ | p{.6\textwidth} | }{\cite{joshi05:generalized} A generalized temporal role-based access control model, 2005 } \\[4ex]

\cline{1-4}
\end{tabularx}
\vline
\label{tab:categorization}
\end{table}

>>>>

\begin{sidewaystable}

\footnotesize
\centering
\caption{Primary sources by classification and properties (Implementation Type / Domain / Evaluation Type(s))}
\vline
\begin{tabularx}{\textwidth}{ | c | c | c | c | l | l | }
\cline{1-4}

@@

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ | p{.95\textwidth} | }{\cite{haibo05:context} A Context-Aware Role-Based Access Control Model for Web Services, 2005 (None / Web Applications / None)} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ | p{.95\textwidth} | }{\cite{jian2008extended} An extended RBAC model based on granular logic, 2008 (None / None / Scenarios)} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{1}{*}{\rotatebox{90}{Const}}}} &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{alam06:constraint} A constraint based role based access control in the SECTET a model-driven approach, 2006 (None / Medical / Scenarios)} \\[4ex] \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{1}{*}{\rotatebox{90}{ Org }}}} &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{zhang06:collaborative} ROBAC: Scalable Role and Organization Based Access Control Models, 2006 (None / Enterprise Workflows / Complexity Analysis)} \\[2.75ex]  \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{2}{*}{\rotatebox{90}{Priv }}}} &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{ni2010privacy} Privacy-aware role-based access control, 2007 (None / Medical / Time-based Performance)} \\[1ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{masoumzadeh2008purbac} PuRBAC: Purpose-Aware Role-Based Access Control, 2008 (None / Web Applications / None)} \\[1ex] \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Task }}}} &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{yao2008task} A Task-Role Based Access Control Model with Multi-Constraints, 2008 (Enterprise / Organizations / Scenarios)} \\ \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{zhou2007team} Team and Task Based RBAC Access Control Model, 2009 (Prototype / Medical / Scenarios)} \\ \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{oh2003task} Task-role-based access control model, 2003 (None / Enterprise Workflows / Scenarios)} \\ \cline{2-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{10}{*}{\rotatebox{90}{ Spatio-Temporal }}}} &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{aich07:STARBAC} STARBAC: Spatiotemporal role based access control, 2007 (None / None / Mathematical Modeling)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{chen08:spatio-temporal} On spatio-temporal constraints and inheritance in role-based access control, 2008 (None / Pervasive Computing / Mathematical Modeling)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{samuel07:spatio-temporal} A framework for specification and verification of generalized spatio-temporal role based access control model, 2007 (None / Medical / Scenarios)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{chandran05:llt} LoT-RBAC: A Location and Time-Based RBAC Model, 2005 (None / Mobile / None)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{ray07:spatio} A Spatio-temporal Role-Based Access Control Model, 2007 (None / Pervasive / Scenarios, Comparison to Standard)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{}  &
\multicolumn{2}{ | p{.9\textwidth} | }{\cite{aich09:role} Role Based Access Control with Spatiotemporal Context for Mobile Applications, 2009 (Enterprise / Medical, Mobile / Complexity Analysis, Time-based performance)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Spatial }}}} &
\multicolumn{1}{ | p{.9\textwidth} | }{\cite{damiani2007geo} GEO-RBAC: a spatially aware RBAC, 2005 (None / Medical / Mathematical Modeling)} \\[0.25ex] \cline{4-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{}  &
\multicolumn{1}{ | p{.9\textwidth} | }{\cite{ray2006lrbac} LRBAC: A location-aware role-based access control model, 2006 (None / Mobile / None)} \\[0.25ex] \cline{4-4}

\multicolumn{1}{ c }{}  & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{}  &
\multicolumn{1}{ | p{.9\textwidth} | }{\cite{hansen2003spatial} Spatial role-based access control model for wireless networks, 2003 (None / Medical / Mathematical Modeling)} \\[0.25ex] \cline{3-4}

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Temp }}}} &
\multicolumn{1}{ | p{.9\textwidth} | }{\cite{joshi05:generalized} A generalized temporal role-based access control model, 2005 (None / Enterprise Workflows / Scenarios)} \\[4ex]

\cline{1-4}
\end{tabularx}
\vline
\label{tab:categorization}
\end{sidewaystable}
